---
title: "Lidar Tutorial"
output: html_notebook
---

This tutorial is an introduction to working with the products of lidar data collection from airborne platforms (e.g., drones) using the statistical software `R`, specifically the library `lidR`.

*Language:* `R`

*Packages:*

| Name | Description | Link |
|:--|:--|:--|
| `lidR` | Package to view and process LiDAR point clouds in `R` | https://cran.r-project.org/web/packages/lidR/index.html |
| `future` | Package to enable parallel processing for more efficient computation | https://cran.r-project.org/web/packages/future/index.html |

## Nomenclature

* *lidar:* Light Detection and Ranging, a method of remote sensing using pulsed lasers to measure distances.
* *LAS:* A file format for storing the 3D point cloud data collected by LiDAR. A related file format is LAZ, which is the compressed version of LAS.
* *Pulse:* A discrete "shot" of laser emission from the LiDAR platform that is used to measure distance from the target surface, which in turn are the basis of point cloud data. 

## Data Details

We will use point cloud data collected over the [U.S. Dairy Forage Research Center Far](https://www.ars.usda.gov/ARSUserFiles/50901500/About%20Us%20and%20Contact%20Information/pdfs/farm%20at%20PdS.pdf) by the [3D Elevation Program](https://www.usgs.gov/3d-elevation-program), a program at the USGS that is conducting nationwide LiDAR surveys.

* Repository: [USGS National Map](https://apps.nationalmap.gov/lidar-explorer/#/)
* Dataset name: The datasets we are using are a subset of the elev3dep project  [WI_8County_2020_A20](https://rockyweb.usgs.gov/vdelivery/Datasets/Staged/Elevation/LPC/Projects/WI_8County_2020_A20/WI_8County_Sauk_2020/browse/)
 * *Note that the data are available in the tutorial folder*

## Anlysis Steps

* Load in LAS data
* Exploring the LAS data structure in `lidR`
  * Attributes
  * Plotting
* Larger geographic coverage with the `lasCatalog` in `lidR`

## Step 0: Load Libraries and data

```{r load_libraries}
library(lidR)
library(future)
library(devtools)
library(ggplot2)
library(tidyverse)

# Here is the directory on Atlas
lasdir <- "/Users/kevinli/Documents/Data/lidar/PdS_3dep/"

# look at the contents of the folder:
lasfiles <- list.files(lasdir, full.names=TRUE)

```

## Step 1: Exploring the LAS data structure in `lidR`

Individual lidar point clouds (lasfiles) are loaded using the function `readLAS`. Read in one of the dataset point clouds using `readLAS`:

```{r readLAS}

las <- readLAS(lasfiles[1])

```

You can plot it in 3D using the `plot` command in `lidR`.

Note that you can specify what attribute to plot using the `color` argument. Try plotting by `Classification`.

```{r plotLAS}

plot(las, color = "Classification")

```

## Step 2: Classifying the ground

Now we will try classifying the ground for this lidar tile. We will read in a new tile ignoring the classification information to pretend it hasn't been done yet. We're going to use the ground classification algorithm `pmf`, which stands for 'progressive morphological filter'.

```{r classify_las}

las_unclass <- readLAS(lasfiles[1], select="xyzrn")

# we need to set up the ground classification parameters

las_gnd <- classify_ground(las_unclass, algorithm = pmf(ws = 5, th = 3))

# view classification results

plot(las_gnd, color = "Classification")

```

To get a better idea of how the classification worked, we can plot a cross section in ggplot. There is a function in `R/plot_cross.r` that can be used for this.

```{r plot_cross}
source("R/plot_cross.r")

plot_cross(las_gnd, transect_len = 300)

```

From the cross section, it looks like we might be classifying some buildings as the ground. We aren't going to worry about this too much for the tutorial but there are ways we could try to address this, like using 

```{r pmf2}
# we will start from the unclassified version of the point cloud again

ws <- c(9, 24, 60, 120)
th <- seq(1, 6, length.out = length(ws))

las_pmf2 <- classify_ground(las_unclass, algorithm = pmf(ws=ws, th=th))

plot_cross(las_pmf2, transect_len = 300)
```

## Step 3: Create a digital terrain model

With the ground identified, we can use this information to create a digital terrain model (DTM). The DTM has a lot of applications, like for modeling hydrology. In our case, we will use the DTM to normalize the point cloud, which allows us to treat the points as if they were collected above a flat surface. This way, the height of the point in the point cloud corresponds to height above the ground in the real world. This will be helpful for modeling vegetation height.

There are also several methods of making a DTM in `lidR`. We will use the 'triangular irregular network' approach, which is a fast and robust method. It uses the Delauney triangulation method, which is simple and requires no parameters.

```{r tin_las}

dtm_tin <- rasterize_terrain(las_pmf2, res = 30, algorithm = tin())

plot_dtm3d(dtm_tin, bg = "white") 

```


## Step 4: Height normalization

Once we have the digital terrain model, we can use it to normalize the heights of the point cloud.

```{r normalization}

nlas <- las_pmf2 - dtm_tin

```

It doesn't look very different because the terrain is very flat here. We can plot a histogram of elevation to check that ground points are now zero. We use the `lidR` function `filter_ground` to only select ground points and plot their Z axis:

```{r ground_z}

hist(filter_ground(las_pmf2)$Z, main = "Not normalized", xlab="Elevation")

hist(filter_ground(nlas)$Z, main = "Normalized", xlab="Elevation")

```

Notice that the ground points are centered at 0 now but they're not all exactly 0. This is because we normalized against the digital terrain model, which generalizes the elevation to the pixel level. The exact elevation at each point within the pixel is likely a little off from the the pixel value.

Another way to normalize the point cloud is to reference the ground points within the point cloud itself. The `normalize_height` function uses a spatial interpolation function to create a continuous surface of ground points to use for normalization rather than the DTM grid. We need to give it an interpolation method, so we use the TIN method again.

```{r point_normalization}

nlas2 <- normalize_height(las_pmf2, tin())

```

If we look at the histogram of ground points from this normalization, we can see they are all at 0.

```{r ground_z_2}

hist(filter_ground(nlas2)$Z, main = "Normalized", xlab="Elevation")

```

## References

Guo, Q., Su, Y., Hu, T., Guan, H., Jin, S., Zhang, J., Zhao, X., Xu, K., Wei, D., Kelly, M., & Coops, N. C. (2021). Lidar Boosts 3D Ecological Observations and Modelings: A Review and Perspective. IEEE Geoscience and Remote Sensing Magazine, 9(1), 232–257. https://doi.org/10.1109/MGRS.2020.3032713

Roussel, J.-R., Goodbody, T. R. H., & Tompalski, P. (2023). The lidR package. https://r-lidar.github.io/lidRbook/



